\chapter*{Acknowledgments}

This journey has been intense, sometimes exhausting, but above all, deeply rewarding.

My sincere thanks go to Michele Loreti, who has been a constant guide since the beginning of my academic path at the University of Camerino. 
%Your support over the years has shaped my growth and brought me to this point.

I am also grateful to Marcello Bonsangue for his kindness, availability, constant smile, and thoughtful feedback.

To the PhD students and staff at LIACS, thank you for creating such a stimulating and welcoming environment. From whiteboard discussions to the many coffee breaks, you made each day more engaging and enjoyable.

To my family, thank you for your unwavering support and for always being there. Mum, you have been my constant reference point for everything, offering practical advice and support whenever I needed it.

To my friends in Camerino, thank you for making that small town feel like the center of the universe. A special thank you to Alice and Francesco, my two anchors during the most overwhelming times.

To Valentijn, thank you for making my time in the Netherlands even more beautiful and for being the reason it truly felt like home. You stood by me during the most life-changing months, bringing peace and calm when I needed it most.

Thank you to everyone I crossed paths with during this journey. Whether you lifted me up or challenged me, you helped shape who I am today.

Finally, I am grateful to the version of myself who started this journey three years ago. You didn't know exactly where you were going, but you kept moving forward, staying curious, brave, and determined. Today, I'm proud of the person you've become. It wasn't easy, but it was worth it.

