\section{Detailed Models of Quantum Finite Automata}
\label{sec:detailed-models}
Among the numerous variants of \glspl{qfa}, the measure-once and measure-many models stand out as two of the most foundational and widely studied frameworks. Respectively known as \gls{mo-1qfa} and \gls{mm-1qfa}, these automata provide a minimalistic yet powerful theoretical playground for investigating the principles of quantum computation with finite memory. Their significance lies not only in their historical development as some of the earliest quantum models for language recognition \cite{moore2000quantum, kondacs1997power}, but also in their role as archetypal examples in the study of quantum-classical computational boundaries.

A \gls{mo-1qfa} performs unitary operations throughout the input reading process and conducts a single measurement only at the end of the computation. This model, introduced by Moore and Crutchfield \cite{moore2000quantum}, can recognise only a restricted class of \glspl{reg}, such as the so-called group languages \cite{brodsky2002characterizations}. In contrast, the \gls{mm-1qfa}, introduced by Kondacs and Watrous \cite{kondacs1997power}, allows measurements after each symbol is processed, enabling it to recognise a strictly larger class of \glspl{reg}, though still not the full class.

The primary importance of \gls{mo-1qfa} and \gls{mm-1qfa} is not just theoretical. These models are particularly suitable for demonstrating techniques for the compilation of quantum automata into quantum circuits, as we will explore in Chapter ~\ref{chap:automata-to-circuits}. Due to their simpler architecture—one-way movement, discrete time steps, and finite-dimensional Hilbert spaces—these automata provide an ideal framework for illustrating how abstract automaton transitions can be implemented using quantum gates and projective measurements. In this thesis, they will serve as canonical models to illustrate the compilation strategy, setting a baseline for comparison with more advanced or generalised \gls{qfa} variants.


\input{QuantumFiniteAutomata/DetailedModels/Models/MOQFA.tex}
\input{QuantumFiniteAutomata/DetailedModels/Models/MMQFA.tex}