\chapter{Background}  

The study of quantum automata theory necessitates a thorough grounding in both classical computational models and the quantum mechanical principles that redefine their capabilities. This chapter systematically establishes the conceptual foundation for analyzing quantum automata by first revisiting classical finite automata—the cornerstone of formal language theory—and then introducing the quantum mechanical framework that enables novel computational paradigms.

We begin with an in-depth exploration of classical finite automata, which serve as the theoretical bedrock for understanding computational limits and language recognition. Deterministic finite automata (DFAs), non-deterministic finite automata (NFAs), probabilistic finite automata (PFAs), and their two-way variants are analyzed through their formal definitions, operational dynamics, and closure properties. These models collectively define the boundaries of classical computation, particularly in recognizing regular languages and their limitations in handling context-free or stochastic languages. The analysis draws on foundational works such as Hopcroft et al. \cite{hopcroft2006introduction}, which formalized the equivalence between DFAs and NFAs, and Rabin's seminal work on probabilistic automata \cite{rabin1963probabilistic}, which expanded the class of recognizable languages through probabilistic acceptance criteria.

The discourse then transitions to quantum mechanical principles essential for quantum computation. Key concepts such as qubit representation, quantum superposition, and entanglement are contextualized within computational frameworks, emphasizing their departure from classical bit-based processing. The measurement postulate and its implications for probabilistic outcomes are discussed in relation to quantum state collapse, a critical distinction from classical probabilistic models. These principles are synthesized with insights from Nielsen and Chuang's definitive text on quantum computation \cite{nielsen2010quantum}, which provides the mathematical formalism for quantum operations.

The chapter's structure is designed to mirror the hierarchical taxonomy developed in later chapters. By first rigorously defining classical models and their limitations, followed by an exposition of quantum principles and their computational implications, the groundwork is laid for analyzing hybrid models such as the one-way quantum finite automaton with classical states (1QFAC) \cite{zheng2012one}. Each section deliberately connects theoretical constructs to practical considerations, such as the role of decoherence in open quantum systems \cite{breuer2002theory} and its impact on automata design. This approach ensures that subsequent discussions of quantum automata variants are rooted in both mathematical rigor and physical realizability.

\input{Background/ClassicalFiniteAutomata/ClassicalFiniteAutomata.tex}
\input{Background/QuantumMechanicsFoundations/QuantumMechanicsFoundations.tex}