\chapter{Background}  
\label{chap:background}

The study of quantum automata theory necessitates a thorough grounding in both classical computational models and the quantum mechanical principles that redefine their capabilities. This chapter systematically establishes the conceptual foundation for analyzing quantum automata by first revisiting classical finite automata—the cornerstone of formal language theory—and then introducing the quantum mechanical framework that underpins novel computational paradigms.

We begin with an in-depth exploration of classical finite automata, which serve as the theoretical bedrock for understanding computational limits and language recognition. \glspl{dfa}, \glspl{nfa}, \glspl{pfa}, and their two-way variants are analysed through their formal definitions, operational dynamics, and closure properties. These models collectively define the boundaries of classical computation, particularly in recognizing regular languages and in delineating limitations when handling context-free or stochastic languages. Foundational works such as Hopcroft et al. \cite{hopcroft2006introduction, sipser2012introduction}—which formalised the equivalence between \glspl{dfa} and \glspl{nfa}—and Rabin's seminal work on probabilistic automata \cite{rabin1963probabilistic} have played a pivotal role in shaping this understanding.

The discussion then transitions to the essential principles of quantum mechanics required for quantum computation. Key concepts such as qubit representation, quantum superposition, entanglement, and the measurement postulate are introduced and contextualised within computational frameworks. These principles fundamentally depart from classical bit-based processing by allowing phenomena such as state interference and probabilistic state collapse. The mathematical formalism provided by Nielsen and Chuang \cite{nielsen2010quantum} serves as the backbone for these quantum operations.

\input{Background/ClassicalFiniteAutomata/Main.tex}
\input{Background/QuantumMechanicsFoundations/Main.tex}
\input{Background/QuantumCircuits/Main.tex}
